---
title: "Starting with with the dplyr, ggplot2 packages, and the pipe operator |>"
author: "Han Olff"
date: "2024-08-29"
format: html
editor: 
  markdown: 
    wrap: 72
editor_options:
  chunk_output_type: console
---

# Working with the dplyr and ggplot2 packages

Load the required libraries :

```{r message = FALSE, warning = FALSE}
# restore your librar; keep at the start of every script to make sure the library version is uptodate
renv::restore()

# clear the R environment
rm(list = ls())

# load the required packages
library(tidyverse)  # this loads a group of related packages, including readr, dplyr and ggplot2
```

We will work with an online database of the Schiermonnikoog transect
study, that you will collect additional data for next week in the field.
In this example, we work with measurements of cockles (a bivalve
mollusc) on their width and length. From the database Epibenthos, we
will work with the table
[FactCockles](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1538766002#gid=1538766002).
See the documentation of the different variables in the table
[MetVariables](https://docs.google.com/spreadsheets/d/1E1vUWAHhse7fhjBf94Kiog3Rsj7IkMuFtP-BTQX2oI8/edit?gid=1290622213#gid=1290622213)

### Read the datafile from the Google sheets database

To read the data in R, you first need to know the published csv link of
the FactCockles table. In the database, you can find this link in the
table MetTables. It is produced in Google Sheets throught the menu
File/Share/Publish to web and then selecting the table and output as
csv.

Read the FactCockleSize table with read_csv, which reads it as tibble
(formatted dataframe)

```{r}
FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv") #instead of read.csv, read_csv makes a special dataframe (a tibble)
print(FactCockleSize)
names(FactCockleSize)
nrow(FactCockleSize)

```

Plot with ggplot the relation between cockle thickness (thickness_mm, as
x) and length (length_mm), showing each year with a different color, and
add a regression line through all the years

```{r}
ggplot(data=FactCockleSize |> dplyr::filter(thickness_mm <300), #filter for outliers
       mapping = aes(x=thickness_mm, 
                     y=length_mm, 
                     col = as.factor(year))) +
  geom_point() +
  geom_smooth(method = 'lm', col = 'orange') # adds a regressionline in the plot

#in klas
ggplot2::ggplot(data=FactCockleSize , # what dataset
                mapping=aes(x=length_mm, y=thickness_mm)) + # what is on the axes
  geom_point() 

#there is a big outlier, likely a data entry mistake, remove it
FactCockleSize |> dplyr::filter(thickness_mm>500) #first show which datapoint is the outlier

FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv") |> dplyr::filter(thickness_mm<500) #filter the original dataset for the outlier, already when initially uploading the datset OR

#filter for this specific observation, this is better 
FactCockleSize <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSpormjGiM4uWMTIZYm9Upt5j1Ige_Wu9CrhGGdvjXwxmHP2R6S7JDyEmiGTn3fXihl5s5yBn_sdo8h/pub?gid=1538766002&single=true&output=csv") |> dplyr::filter(CockleObs_ID!=1531, CockleObs_ID!=469) |>
  dplyr::mutate(year=factor(year)) #make year a factor, because it is written down numerically while being a factor

print(FactCockleSize)

#again make the plot
ggplot2::ggplot(data=FactCockleSize , # what dataset
                mapping=aes(x=length_mm, y=thickness_mm)) + # what is on the axes
  geom_point() 

#find that other outlier
FactCockleSize |> dplyr::filter(thickness_mm>10 & length_mm<5)
#also remove observation 469 above
```

Further explore the plot with a regression line through all the data

```{r}
#plot with thickness_mm per year
ggplot(data=FactCockleSize |> dplyr::filter(thickness_mm <300), #filter for outliers
       mapping = aes(x=year, 
                     y=thickness_mm)) +
  geom_point() +
  geom_smooth(method = 'lm', col = 'orange') # adds a regressionline in the plot

#plot with length_mm per year
ggplot(data=FactCockleSize, #filter for outliers
       mapping = aes(x=year, 
                     y=length_mm, 
                     col = year)) +
  geom_point() +
  geom_smooth(method = 'lm', col = 'orange') # adds a regressionline in the plot


#in klas
#plot the data with a regression line
ggplot2::ggplot(data=FactCockleSize , # what dataset
                mapping=aes(x=length_mm, y=thickness_mm)) + # what is on the axes
  geom_point() +
  geom_smooth(method = 'lm')

#calculate a linear regression
model_lm<-lm(thickness_mm~length_mm, data=FactCockleSize)
summary(model_lm)

#equation: thickness_mm = -0.353442 + 0.699362 * length_mm
#a cockle of length 20 mm has an predicted width of:
-0.353442 + 0.699362 * 20
# this regression is significant, R2=0.90, P<0.001 (***)
```

Make same plot but showing a separate regression line per year

```{r}
# color the points by year, but plot one regression line
ggplot(data=FactCockleSize |> 
        dplyr::filter(thickness_mm <300), #filter for outliers
       mapping = aes(x=thickness_mm, 
                     y=length_mm,
                     col = as.factor(year))) +
  geom_point() +
  geom_smooth(method = 'lm') # adds a regressionline in the plot


#in klas
ggplot2::ggplot(data=FactCockleSize , # what dataset
                mapping=aes(x=length_mm, y=thickness_mm)) + # what is on the axes
  geom_point(mapping=aes(col = year)) +
  geom_smooth(method = "lm", col = 'black')
```

Make a panel plot where with each year is shown as a separate graph

```{r}
ggplot(data=FactCockleSize |> 
        dplyr::filter(thickness_mm <300), #filter for outliers
       mapping = aes(x=thickness_mm, 
                     y=length_mm,
                     col = as.factor(year))) +
  geom_point() +
  facet_wrap(~ year) +
  geom_smooth(method = 'lm', col = 'black')


#in klas
drop_na(FactCockleSize) |>   # drops any rows that are missing values
 # dplyr::filter(year %in% c(2014,2015,2017,2018,2020,2021)) |>  #filter for particular years
 # dplyr::mutate(year=factor(year))  |>
    ggplot2::ggplot(mapping=aes(x=length_mm, y=thickness_mm)) + # what is on the axes
      geom_point(mapping=aes(col = year)) +
      xlab("cockle length (mm)") +
      ylab("cockle width (mm)") +
      geom_smooth(method = "lm", col = 'black') +
      facet_wrap(~year) #adds a panel plot per year
```

We conclude from this that: \* there were two important outliers in the
dataset that were removed after visual inspection \* the regression
between length and width is abou the same for every year, we safely can
use only length as a proxy for the biomass of an individual cockle
